---
title: "seaPiper usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seaPiper usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

seaPiper is a user interface to studying the results of differential gene
expression analysis. Note that it does not allow any analysis per se. This
is an important distinction: the analysis results are generated using a
pipeline – that allows them to be reproducible and well documented; every
analysis result can be traced back to the code that generated it, versions
of packages used to generate it etc. Analyses performed interactively (in a
graphical user interface), for example in a browser, are notoriously hard
to perform in a way that is accountable and reproducible.

# Building `data` for `seapiper()`

The `seapiper()` function expects a `seaPiperData` object (`data` argument).
There are now three supported ways to create it:

1. From a loaded `Rseasnap` pipeline:

```{r eval=FALSE}
pip <- Rseasnap::load_de_pipeline(config_file = "DE_config.yaml")
data <- seapiperdata_from_rseasnap(pip)
seapiper(data)
```

2. From in-memory objects:

```{r eval=FALSE}
data <- seapiperdata_from_objects(
  cntr = cntr,
  annot = annot,
  exprs = exprs,
  primary_id = "PrimaryID"
)
seapiper(data)
```

3. From a custom YAML file with RDS object paths:

```{r eval=FALSE}
data <- seapiperdata_from_yaml("seapiper_data.yaml")
seapiper(data)
```

Example `seapiper_data.yaml`:

```yaml
datasets:
  default:
    annot: annot.rds
    cntr: cntr.rds
    tmod_res: tmod_res.rds
    tmod_dbs: tmod_dbs.rds
    tmod_map: tmod_map.rds
    tmod_gl: tmod_gl.rds
    config: config.rds
    covar: covar.rds
    rld: rld.rds
```

Expected object types for each key:

- `annot`: `data.frame` with gene annotation (ideally with `PrimaryID`).
- `cntr`: named `list` of contrast `data.frame`s.
- `tmod_res`: named `list` of tmod results per contrast.
- `tmod_dbs`: named `list` of tmod DB objects (or wrappers with `$dbobj`).
- `tmod_map`: tmod mapping object (`get_tmod_mapping` output).
- `tmod_gl`: nested gene-ranking object (same shape as `gl.rds`).
- `config`: sea-snap style configuration `list`.
- `covar`: sample covariates `data.frame` (required).
- `rld`: transformed expression matrix/data.frame (genes x samples).
- `pca`: PCA matrix/data.frame (samples x PCs, optional).
